A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2022-02-01 analysis/paper/paper.Rmd
Wokring on Brunlanes sites

The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways. The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [but see @mjærum2018; @gundersen2013; @schulke2020]. This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2021; @prescott2020]. associated involving a change agro-pastoralism, the farm,  means that the site locations are now more withdrawn from the shoreline [@bakka1971; @prescott2012]. That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods. At an earlier stage, at the transition to the Early Neolithic (c. 3900 BCE), pottery is introduced to the sites, and there are some possible indications of an initial, small-scale uptake of agriculture at some sites in the Oslo fjord region. However, this initial uptake appears to be relatively small-scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, followed by a complete de-Neolithisation in the Middle Neolithic. [@nielsen2021] has recently argued that while there, the initial uptake of in the Early Neolithic is combined with a more complex settlement pattern, and argues that the simple forager/farming dichotomy has underplayed the variation present in the Early and Middle Neolithic data. The empirical expectation for the above outlined development would thus be a more varied relation between sites and the shore-line from around 3900 BCE followed by a more decisive shift around 2400 BCE.

@@

One of the more comprehensive evaluations of this relationship was done by Solheim and colleagues (@breivik2018; @solheim2020), who compared 102 ^14^C-dates from 29 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area. They found an overlap between the probability distribution of the radiocarbon dates with the shoreline displacement curve for 86.5% of the sites. However, where there was a discrepancy, the main occupation of the sites are still believed to have been shore-bound rather than associated with the deviating ^14^C-dates. This is based on typological and technological characteristics of the assemblages. Whether these mismatches represent later shorter visits that are responsible for the younger radiocarbon dates, or whether these are entirely erroneous results of contamination can be difficult to evaluate. Here this will be entirely dependent on, and follow the discretion of the original excavation reports.

Other previous evaluations of the correspondence between dates informed by radiocarbon dates and shoreline dating have typically followed the same structure as that of @breivik2018, involving a visual inspection of radiocarbon probability distributions plotted against local shoreline displacement curves based on the elevation of the site [e.g. @åstveit2018; @solheim2020]. This approach has a couple of limitations. First of all, the displacement curves are commonly applied directly to larger study areas, with only a few studies having taken the variable uplift-rates into account when performing this comparison [e.g. @fossum2020; @møller1987; @persson2008]. Secondly, with this method, the wider the uncertainty range associated with either radiocarbon date or displacement curve, the higher the probability that the confidence intervals overlap, and the higher the probability that we conclude in favour of our hypothesis. This thus leads to an inferential framework that favours uncertainty, which is hardly desirable. In statistical terms this follows from the fact that while one cannot conclude that two dates are different if their confidence intervals overlap, this does not necessarily mean that they are the same. The question thus necessitates a flip from a null-hypothesis of no significant difference, to one of equivalence [e.g. @lakens2018], as the question of interest is effectively one of synchroneity between events [cf. @parnell2008]. Another limitation of this often-employed method is that it only takes into account the vertical distance between the sites and the sea-level. While this is the main parameter of interest for shoreline dating, the practical implications of a vertical difference in RSL will be highly dependent on local topography and bathymetry. RSL-change can have more dramatic consequences in a landscape characterised by a low relief, as the horizontal displacement of the shoreline will be greater. Taking the spatial nature of the relationship between site and shoreline into account will consequently help get more directly at the behavioural dimension of this relation, and move the analysis beyond a purely instrumental consideration of the applicability of shoreline dating. Suggested ways to help mitigate and integrate the issues presented above into to the analysis are presented in the methods section.

@@

Three of the sites have been associated with farming activities, either in directly or in the form building structures. The first is Nordby 1, which has . Furthermore, the Middle Neolithic phase at Kvastad A2 and Late Neolithic phase at Nauen A are both related to farming. Both of these sites also have ^14^C-dates and lithic inventory associated with Mesolithic forager activities. Following from the expected deviance from the settlement patterns that characterise forager sites, these agricultural phases are thus treated seperately below. Finally, @nielsen2021 has recently suggested that Early and Middle Neolithic features from the otherwise younger sites Bratberg and Lang could be related to early agricultural activity in the Oslo fjord region. Due to the uncertain and somewhat speculative nature of thi suggestion, these are omitted here.

@@

The elevation data used for the analysis is a digital terrain model (DTM) freely available from the Norwegian Mapping Authority [@mapping2019]. After evaluating the options I opted for the 10m resolution DTM rather than the higher-resolution 1m version. In addition to resulting in considerably less processing time, the higher resolution elevation model is more vulnerable to smaller-scale modern disturbances that the 10m version is not impacted by. The 10m DTM is a down-sampled version of the 1m resolution DTM, which is based on aerial laser scanning using a minimum of two and up to five points per m^2^ [@mapping2019].

@@

Not least following from the fact that very few pre-Boreal ^14^C-dates associated with anthropogenic activity have been achieved in Norway, the shoreline dating of the earliest sites is essential for understanding the pioneer settlement and the initial colonisation of the Scandinavian peninsula [e.g. @bang-andersen2012; @breivik2014; @glorstad2016]. The shoreline dated pre-Boreal sites from the Brunlanes-project are among the earliest known sites in Norway [@jaksland2014a; @jaksland2012; @jaksland2012a]. These have a distinct Early Mesolithic artefact inventory and are situated in a steep area of the landscape where it would be difficult to envision any use of the sites after the sea retreated any significant distance from their location. In the original publication of the sites, @jaksland2014 provides a thorough discussion of shoreline dating in general, and as used for the dating of the Brunlanes sites specifically. A comparison of his results and the ones achieved using the above-outlined approach are given in Figure. These are largely in agreement. The sites are located more or less on the isobase of the displacement curve for Larvik, and negligible adjustments were therefore needed. The small discrepancies between the results mainly follow from the fact that a slightly updated version of the displacement curve is applied here [cf. @sorenseninpress]. Jaksland's dates are given a flat 200 and 50 year uncertainty range starting from what he gives as the earliest possible date. The 200 year uncertainty range is given if the sites are considered in isolation, while the argument for the uncertainty range of only 50 years is based on the location of the sites relative to each other. Since they are located in such a constrained and steep area of the landscape, the difference in elevation between the sites is argued to establish their relative date and thus constrain the uncertainty range so that they don't overlap between sites located at different elevations. This information is not integrated in the approach outlined here, but could justify reducing the upper bound of the uncertainty ranges, especially for Pauler 1 and Pauler 2. While the accuracy is of course ultimately dependent on the veracity of the geological reconstruction, the steepness of the curve does nonetheless provide very precise dates, and the agreement with Jakslands thorough consideration does speak further to the general applicability of the approach to shoreline dating outlined in this study. This example also highlights the fact that a consideration of the surrounding terrain and other sites can help in increasing the precision of the method if this can be used to exclude certain sea-levels as unlikely for when a site was in use. One approach to this could also be to assess the spatial implication of a proposed shoreline date by simulating the adjusted sea-levels, as is done for Pauler 1 in Figure, followed for example by a visual evaluation of the topography or by measuring the steepness of the slope to the shoreline. Based on this, it could be possible to exclude certain elevations as unlikely for the position of the shoreline. Such approaches would make less of an impact in this setting, where the 95% probability range is already quite constrained, but could considerably improve the precision of the method in cases where the displacement curve is flatter (cf. Figure \@ref(fig:shoredate)).

```{r brunlanes, echo = FALSE, fig.cap = "Shoreline dating the Brunlanes sites. A) The result of applying the approach to shoreline dating outlined here, plotted against the dates provided by Jaksland (2014) given in black. B) Map of the Pauler sites and Sky 1 with simulated sea-level using the probability density associated with the shoreline date for Pauler 1.", out.width = "525px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/brunlanes.png"))
```

>>>>

The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways. The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brø but see @mjærum2018; @gundersen2013; @schulke2020]. This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2021; @prescott2020]. associated involving a change agro-pastoralism, the farm,  means that the site locations are now more withdrawn from the shoreline [@bakka1971; @prescott2012]. That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods. At an earlier stage, at the transition to the Early Neolithic (c. 3900 BCE), pottery is introduced to the sites, and there are some possible indications of an initial uptake of agriculture at some sites in the Oslo fjord region. However, this appears to be small-scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, followed by a complete de-Neolithisation in the Middle Neolithic [@nielsen2019; @hinsch1955]. [@nielsen2021b] has recently argued that while there, the initial uptake of agriculture in the Early Neolithic is combined with a more complex settlement pattern, and that the simple forager/farming dichotomy has underplayed the variation present in the Early and Middle Neolithic data. Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a more varied association between sites and the shore-line from around 3900 BCE, followed by a more decisive shift around 2400 BCE.

@@

One of the more comprehensive evaluations of this relationship was done by Solheim and colleagues (@breivik2018; @solheim2020), who compared 102 ^14^C-dates from 29 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area. They found an overlap between the probability density of the radiocarbon dates with the shoreline displacement curve for 86.5% of the sites. However, where there was a discrepancy, the main occupation of the sites are still believed to have been shore-bound rather than associated with the deviating ^14^C-dates. This is based on typological and technological characteristics of the assemblages. Whether these mismatches represent later shorter visits that are responsible for the younger radiocarbon dates, or whether these are entirely erroneous results of contamination can be difficult to evaluate. Here this will be entirely dependent on, and follow the discretion of the original excavation reports.

Other previous evaluations of the correspondence between dates informed by radiocarbon dates and shoreline dating have typically followed the same structure as that of @breivik2018, involving a visual inspection of radiocarbon probability density functions plotted against local shoreline displacement curves based on the elevation of the site [e.g. @åstveit2018; @solheim2020]. This approach has a couple of limitations. First of all, the displacement curves are commonly applied directly to larger study areas, with only a few studies having taken the variable uplift-rates into account when performing this comparison [e.g. @fossum2020; @møller1987; @persson2008]. Secondly, with this method, the wider the uncertainty range associated with either radiocarbon date or displacement curve, the higher the probability that the confidence intervals overlap, and the higher the probability that we conclude in favour of our hypothesis. This thus leads to an inferential framework that favours uncertainty, which is hardly desirable. In statistical terms this follows from the fact that while one cannot conclude that two dates are different if their confidence intervals overlap, this does not necessarily mean that they are the same. The question thus necessitates a flip from a null-hypothesis of no significant difference, to one of equivalence [e.g. @lakens2018], as the question of interest is effectively one of synchroneity between events [cf. @parnell2008]. Another limitation of this often-employed method is that it only takes into account the vertical distance between the sites and the sea-level. While this is the main parameter of interest for shoreline dating, the practical implications of a vertical difference in RSL will be highly dependent on local topography and bathymetry. RSL-change can have more dramatic consequences in a landscape characterised by a low relief, as the horizontal displacement of the shoreline will be greater. Taking the spatial nature of the relationship between site and shoreline into account will consequently help get more directly at the behavioural dimension of this relation, and move the analysis beyond a purely instrumental consideration of the applicability of shoreline dating. Suggested ways to help mitigate and integrate the issues presented above into to the analysis are presented in the methods section.

@@

Three of the sites have been associated with farming activities, either in directly or in the form building structures. The first is Nordby 1, at which the ^14^C-dates are associated with a Late Neolithic long-house. The Middle Neolithic phase at Kvastad A2 and Late Neolithic phase at Nauen A are both related to farming. Both of these sites also have radiocarbon dates and lithic inventory associated with Mesolithic forager activities. Following from the expected deviance from the settlement patterns that are to characterise forager sites, these agricultural phases are thus treated separately below. Finally, @nielsen2021b has recently suggested that Early and Middle Neolithic features from the otherwise younger sites Bratsberg and Larønningen could be related to early agricultural activity in the Oslo fjord region. Due to the uncertain and somewhat speculative nature of this suggestion, these are omitted here.

@@

The elevation data used for the analysis is a digital terrain model (DTM) freely available from the Norwegian Mapping Authority [@mapping2019]. It was here opted for the 10m resolution DTM rather than the higher-resolution 1m version. In addition to resulting in considerably less processing time, the higher resolution elevation model is more vulnerable to smaller-scale modern disturbances that the 10m version is not impacted by. The 10m DTM is a down-sampled version of the 1m resolution DTM, which is based on aerial laser scanning using a minimum of two and up to five points per m^2^ and has a vertical error of  [@mapping2019].

@@

Not least following from the fact that very few pre-Boreal ^14^C-dates associated with anthropogenic activity have been achieved in Norway, the shoreline dating of the earliest sites is essential for understanding the pioneer settlement and the initial colonisation of the Scandinavian peninsula [e.g. @bang-andersen2012; @breivik2014; @glorstad2016]. The shoreline dated pre-Boreal sites from the Brunlanes-project are among the earliest known sites in Norway [@jaksland2014a; @jaksland2012; @jaksland2012a]. These have a distinct Early Mesolithic artefact inventory and are situated in a steep area of the landscape where it would be difficult to envision any use of the sites after the sea retreated any significant distance from their location. In the original publication of the sites, @jaksland2014 provides a thorough discussion of shoreline dating in general, and as used for the dating of the Brunlanes sites specifically. A comparison of his results and the ones achieved using the above-outlined approach are given in Figure Figure \@ref(fig:brunlanes)A. These are largely in agreement. The sites are located more or less on the isobase of the displacement curve for Larvik, and negligible adjustments were therefore needed. The small discrepancies between the results mainly follow from the fact that a slightly updated version of the displacement curve is applied here [cf. @sorenseninpress]. Jaksland's dates are given a flat 200 and 50 year uncertainty range starting from what he gives as the earliest possible date. The 200 year uncertainty range is given if the sites were to be considered in isolation, while the argument for the uncertainty range of only 50 years is based on the location of the sites relative to each other. Since they are located in such a constrained and steep area of the landscape, the difference in elevation between the sites is argued to establish their relative date and thus constrain the uncertainty ranges so that they don't overlap. This information is not integrated in the approach outlined here, but could justify reducing the the uncertainty ranges, especially for Pauler 1 and Pauler 2. While the accuracy is of course ultimately dependent on the veracity of the geological reconstruction, the steepness of the curve does nonetheless provide very precise dates. This example also highlights the fact that a consideration of the surrounding terrain and other sites can help in increasing the precision of the method if this can be used to exclude certain sea-levels as unlikely for when a site was in use. One approach to this could also be to assess the spatial implication of a proposed shoreline date by simulating the adjusted sea-levels, as is done for Pauler 1 in Figure \@ref(fig:brunlanes)B, followed for example by a visual evaluation of the topography or by measuring the steepness of the slope to the shoreline. Based on this, it could be possible to exclude certain elevations as unlikely for the position of the shoreline. Such approaches would make less of an impact in this setting, where the 95% probability range is already quite constrained, but could considerably improve the precision of the method in cases where the displacement curve is flatter (cf. Figure \@ref(fig:shoredate)).

```{r brunlanes, echo = FALSE, fig.cap = "Shoreline dating of the Brunlanes sites using elevations provided by Jaksland (2014:tab.4). A) The result of applying the approach to shoreline dating outlined here. Dates provided by Jaksland (2014) are plotted in black starting from the earliest possible date and where the box indicates a 50 year uncertainty range, and the line 200 years. B) Map of the Pauler sites and Sky 1. The sea-level has been simulated using the probability density associated with the shoreline date for Pauler 1 given as the yellow point (see also map in Jaksland 2014:fig.12a).", out.width = "525px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/brunlanes.png"))
```
